---
title: "CyclisticCaseStudy"
author: "Wyatt Humphrey"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)

```
## Load Data
```{r data}
q1_2019_raw <- readr::read_csv("./QuaterlyData/Divvy_Trips_2019_Q1.csv")
q1_2020_raw <- readr::read_csv("./QuaterlyData/Divvy_Trips_2020_Q1.csv")
```

## View Data
```{r}
glimpse(q1_2019_raw)
glimpse(q1_2020_raw)

head(q1_2019_raw)
head(q1_2020_raw)

colnames(q1_2019_raw)
colnames(q1_2020_raw)

```

## Wrangle Data
# Rename Columns
```{r 2019}
q1_2019 <- q1_2019_raw %>% 
  rename(
    ride_id = trip_id,
    rideable_type = bikeid,
    started_at = start_time,
    ended_at = end_time,
    start_station_name = from_station_name,
    end_station_name = to_station_name,
    start_station_id = from_station_id,
    end_station_id = to_station_id,
    member_casual = usertype
  ) %>% 
  mutate(
    ride_id = as.character(ride_id),
    rideable_type = as.character(rideable_type)
  )
  
q1_2020 <- q1_2020_raw
```
# Inspect the data frames
```{r}
str(q1_2019)
str(q1_2020)

```
# Combine quaters into one data frame
```{r}
all_trips <- bind_rows(q1_2019, q1_2020)#, q3_2019)#, q4_2019, q1_2020)

```
# Remove lat, long, birthyear, and gender fields as data was dropped in 2020
```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender,  "tripduration"))

```

## Clean Data
# Inspect Data
```{r}
colnames(all_trips)
nrow(all_trips)
dim(all_trips)
head(all_trips)
str(all_trips)
summary(all_trips)
```

```{r}
table(all_trips$member_casual)
```
# Reassign `member_casual` to correct values
```{r}
all_trips <- all_trips %>% 
  mutate(
    member_casual=recode(member_casual,
                         "Subscriber" = "member",
                         "Customer" = "casual")
  )
```
Check if encoding was correct
```{r}
table(all_trips$member_casual)
```
# Add columns for `date`, `month`, `day`, `year`, and `day_of_week`
```{r}
all_trips$date <- as.Date(all_trips$started_at)
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

# Add `ride_length` column
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

str(all_trips )
```
# Convert `ride_length` to numeric
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length<-as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)
```
# Remove Bad data "HQ QR"
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]

```

## Analysis
# `ride_length` analysis
```{r}
summary(all_trips_v2$ride_length)

```
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```
# See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
# Notice that the days of the week are out of order. Let's fix that.
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```
# analyze ridership data by type and weekday
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()							#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 		# calculates the average duration
  arrange(member_casual, weekday)		
```
# Visualize
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")

```
# Visualization for average duratio
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")

```

# Save as csv

counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
out_dir <- choose.dir()
out_file <- file.path(out_dir, "avg_ride_length.csv") 

readr::write_csv(counts, out_file)

out_file<-file.path(out_dir, "all_trips_v2.csv")
readr::write_csv(all_trips_v2, out_file)

out_file<-file.path(out_dir, ".csv")
readr::write_csv(all_trips, out_file)



